Characteristic polynomials of p-adic matrices

### 2017-02-02 charpoly/charpoly.tex
Working on Section 2

\numberofauthors{3}
\author{
\alignauthor Xavier Caruso\\
  \affaddr{Universit\'e Rennes 1}\\
  \affaddr{\textsf{xavier.caruso@normalesup.org}}
\alignauthor David Roe\\
  \affaddr{Pittsburg University}\\
  \affaddr{\textsf{roed@pitt.edu}}
\alignauthor Tristan Vaccon\\
  \affaddr{Universit\'e de Limoges}\\
  \affaddr{\textsf{tristan.vaccon@unilim.fr}}
}

@@

Let $\chi : M_n(K) \to K_n[X]$ be the characteristic polynomial map.
Our analysis of the precision behavior of $\chi$ rests upon
the computation of its derivative $d\chi$, using \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}.
For a matrix $M \in M_n(K)$ we identify the tangent space $V$
at $M$ with $M_n(K)$ itself, and the tangent space $W$ at $\chi(M)$ with
the space $K_{<n}[X]$ of polynomials of degree less than $n$.
Let $\com(M)$ denote the comatrix of $M$ (when $M \in \GL_n(K)$,
we have $\com(M) = \det(M) M^{-1}$).  Recall
\citelist{\cite{caruso-roe-vaccon:14a}*{Appendix B} \cite{caruso-roe-vaccon:15a}*{\S 3.3}}
that $d\chi(M)$ is given by
\begin{equation} \label{eq:dchi}
d\chi(M): dM \mapsto \tr(\com(X-M) \cdot dM).
\end{equation}

\begin{prop}
\label{prop:surjectivity}
For $M \in M_n(K)$, the following conditions are equivalent:
\begin{enumerate}[(i)]
\renewcommand{\itemsep}{0pt}
\item \label{pt:surj} the differential $d\chi(M)$ is surjective
\item \label{pt:cyc} the matrix $M$ has a cyclic vector (\emph{i.e.} $M$ is similar
to a companion matrix)
\item \label{pt:1d} the eigenspaces of $M$ over the algebraic
closure $\bar{K}$ of $K$ all have dimension $1$
%\item the Jordan blocks of $M$ all have distinct generalized eigenvalues
\end{enumerate}
\end{prop}

@@

For any $A \in \GL_n(K)$, the image of $d\chi(M)$ will be the same
as the image of $d\chi(AMA^{-1})$, so we may assume that
$M$ is a companion matrix.  For a companion matrix, the bottom row of
$\com(X-M)$ consists of $1, X, X^2, \dots, X^{n-1}$ so $d\chi(M)$ is surjective.

To show the inverse, suppose that $M$ has a repeated eigenvalue $\lambda$ over $\bar{K}$.
After conjugating into Jordan normal form over $\bar{K}$, the entries of $\com(X-M)$
will also be block diagonal, and divisible within each block by the product of $(X-\mu)$
for each other eigenvalue $\mu$.  Since $\lambda$ occurs twice as an eigenvalue,
$X - \lambda$ will divide every entry of $\com(X-M)$ and $d\chi(M)$ will not be surjective.
\end{proof}

@@

\begin{prop} \label{prop:mainlem}
Suppose that $M \in M_n(K)$, all the Jordan blocks of $M$ have distinct generalized eigenvalues, and
let $C = \max(\lvert \det(M)^{-1} \rvert, 1)$.  \todo{What if $\det(M) = 0$?}. Then, for all $\rho \in (0, 1]$ and all
$r \in (0, C \cdot \rho^{-1})$, any lattice $H$ such that $B_V^-(\rho r) \subset H \subset B_V(r)$
satisfies:
\begin{equation}
\chi(M + H) = \chi(M) + d\chi(H).
\end{equation}
\end{prop}

@@

@@

\begin{prop}
Let $A$ be the subring of $K[X]$ consisting of polynomials $P$ for
which $P(M) \in M_n(\OK)$. Then $d \chi_M \big(M_n(\OK)\big)$ 
(considered as a submodule of $K[X]/\chi_M(X)$) is stable under 
multiplication by the elements of $A$.
\end{prop}
\begin{proof}
\todo{The image is generated by the entries of $C = \com(X-M)$
and we have $P(M) \cdot C \equiv  P(X) \cdot C \pmod{\chi_M(X)}$ since the entries of $P(X) \cdot C$
are linear combinations with integer coefficients of the entries of $C$.  $C*(X-M) = \chi(M)$ (since matrix * comatrix is determinant).}
\end{proof}

\begin{cor}
If the coefficients of $M$ lie in $\OK$, then $d \chi_M \big(M_n(\OK)\big)$ 
is stable under multiplication by $X$ and hence is a module over $\OK[X]$.
\end{cor}

We can also give a criterion on $M$ that allows an easy description of the image of flat
precision lattices under $d\chi(M)$.
\begin{prop}
Suppose that $M \in M_n(\OK)$, and that the reduction $\bar{M}$ of $M$ modulo $\pi$ has
a cyclic vector.  Then for any $r > 0$, we have $d\chi(M)(B_V(r)) = B_W(r)$.
\end{prop}
\begin{proof}
First note that it suffices to prove the claim for $r=1$ since $d\chi(M)$ is linear.
Next, it suffices to check that $d\chi(M)$ is surjective modulo $\pi$.
Surjectivity of differential
can be seen: write matrix in Jordan normal form.  Comatrix(X - M) surjective iff when you write it
in Frobenius normal form it has only one block.
\end{proof}

>>>>

\numberofauthors{3}
\author{
\alignauthor Xavier Caruso\\
  \affaddr{Universit\'e Rennes 1}\\
  \affaddr{\textsf{xavier.caruso@normalesup.org}}
\alignauthor David Roe\\
  \affaddr{University of Pittsburg}\\
  \affaddr{\textsf{roed@pitt.edu}}
\alignauthor Tristan Vaccon\\
  \affaddr{Universit\'e de Limoges}\\
  \affaddr{\textsf{tristan.vaccon@unilim.fr}}
}

@@

Let $\chi : M_n(K) \to K_n[X]$ be the characteristic polynomial map.
Our analysis of the precision behavior of $\chi$ rests upon
the computation of its derivative $d\chi$, using \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}.
For a matrix $M \in M_n(K)$ we identify the tangent space $V$
at $M$ with $M_n(K)$ itself, and the tangent space $W$ at $\chi(M)$ with
the space $K_{<n}[X]$ of polynomials of degree less than $n$.
Let $\com(M)$ denote the comatrix of $M$ (when $M \in \GL_n(K)$,
we have $\com(M) = \det(M) M^{-1}$) and $d\chi_M$ the differential at $M$.  Recall
\citelist{\cite{caruso-roe-vaccon:14a}*{Appendix B} \cite{caruso-roe-vaccon:15a}*{\S 3.3}}
that $d\chi_M$ is given by
\begin{equation} \label{eq:dchi}
d\chi_M: dM \mapsto \tr(\com(X-M) \cdot dM).
\end{equation}

\begin{prop}
\label{prop:surjectivity}
For $M \in M_n(K)$, the following conditions are equivalent:
\begin{enumerate}[(i)]
\renewcommand{\itemsep}{0pt}
\item \label{pt:surj} the differential $d\chi_M$ is surjective
\item \label{pt:cyc} the matrix $M$ has a cyclic vector (\emph{i.e.} $M$ is similar
to a companion matrix)
\item \label{pt:1d} the eigenspaces of $M$ over the algebraic
closure $\bar{K}$ of $K$ all have dimension $1$
%\item the Jordan blocks of $M$ all have distinct generalized eigenvalues
\end{enumerate}
\end{prop}

@@

For any $A \in \GL_n(K)$, the image of $d\chi$ at $M$ will be the same
as the image of $d\chi$ at $AMA^{-1}$, so we may assume that
$M$ is a companion matrix.  For a companion matrix, the bottom row of
$\com(X-M)$ consists of $1, X, X^2, \dots, X^{n-1}$ so $d\chi_M$ is surjective.

To show the inverse, suppose that $M$ has a repeated eigenvalue $\lambda$ over $\bar{K}$.
After conjugating into Jordan normal form over $\bar{K}$, the entries of $\com(X-M)$
will also be block diagonal, and divisible within each block by the product of $(X-\mu)$
for each other eigenvalue $\mu$.  Since $\lambda$ occurs twice as an eigenvalue,
$X - \lambda$ will divide every entry of $\com(X-M)$ and $d\chi_M$ will not be surjective.
\end{proof}

We also have an analogue of Proposition \ref{prop:surjectivity} for integral matrices.

\begin{prop}
\label{prop:intsurj}
For $M \in M_n(\OK)$, the following conditions are equivalent:
\begin{enumerate}[(i)]
\renewcommand{\itemsep}{0pt}
\item \label{pt:surjp} the image of $M_n(\OK)$ under $d\chi_M$ is $\OK[X] \cap K_{<n}[X]$.
\item \label{pt:cycp} the reduction of $M$ modulo $\pi$ has a cyclic vector
\end{enumerate}
\end{prop}
\begin{proof}
The condition \eqref{pt:surjp} is equivalent to the surjectivity of $d\chi_M$ modulo $\pi$.  The equivalence with
\eqref{pt:cycp} follows the same argument as Proposition \ref{prop:surjectivity}, but over $\OK / \pi\OK$.
\end{proof}

@@

\begin{prop} \label{prop:mainlem}
Suppose that $M \in M_n(K)$ satisfies one of the conditions of Proposition \ref{prop:surjectivity}, and
let $C = \max(\lvert \det(M)^{-1} \rvert, 1)$.  \todo{What if $\det(M) = 0$?}. Then, for all $\rho \in (0, 1]$ and all
$r \in (0, C \cdot \rho^{-1})$, any lattice $H$ such that $B_V^-(\rho r) \subset H \subset B_V(r)$
satisfies:
\begin{equation}
\chi(M + H) = \chi(M) + d\chi_M(H).
\end{equation}
\end{prop}

@@

The relationship between precision and the images of lattices under $d\chi_M$ allows us to
apply Proposition \ref{prop:intsurj} to determine when the precision of the characteristic polynomial
is the minimum possible.

\begin{cor}
Suppose that $M \in \GL_n(\OK)$ is known with precision $O(p^N)$.
Then the characteristic polynomial of $M$ has precision larger than $O(p^N)$
if and only if the reduction of $M$ modulo $\pi$ does not have a cyclic vector.
\end{cor}

\subsection{Stability under multiplication by $X$}

@@

\begin{prop}
Let $A$ be the subring of $K[X]$ consisting of polynomials $P$ for
which $P(M) \in M_n(\OK)$, and $V = d \chi_M \big(M_n(\OK)\big)$
as a submodule of $K[X]/\chi_M(X)$.  Then $V$ is stable under 
multiplication by $A$.
\end{prop}
\begin{proof}
Let $C = \com(X-M)$ and $P \in A$.  By \eqref{eq:dchi}, $V$ is given by
the $\OK$-span of the entries of $C$.   Using the fact that the product of matrix
with its comatrix is the determinant, $(X - M) \cdot C = \chi(M)$ and thus
$P(X) \cdot C \equiv P(M) \cdot C \pmod{\chi_M(X)}$.  The span of the entries
of the left hand side is precisely $P(X) \cdot V$, while the span of the entries
of the right hand side is contained within $V$ since $P(M) \in M_n(\OK)$.
\end{proof}

\begin{cor}
If $M \in M_n(\OK)$, then $d \chi_M \big(M_n(\OK)\big)$ 
is stable under multiplication by $X$ and hence is a module over $\OK[X]$.
\end{cor}

We can also give a criterion on $M$ that allows an easy description of the image of flat
precision lattices under $d\chi(M)$.
\begin{prop}
Suppose that $M \in M_n(\OK)$, and that the reduction $\bar{M}$ of $M$ modulo $\pi$ has
a cyclic vector.  Then $d\chi_M\big(M_n(\OK)\big) = \OK[X] / \chi(M)$.
\end{prop}
\begin{proof}
It suffices to check that $d\chi_M$ is surjective modulo $\pi$.
Surjectivity of differential
can be seen: write matrix in Jordan normal form.  Comatrix(X - M) surjective iff when you write it
in Frobenius normal form it has only one block.
\end{proof}
